---
title: "Structured Variance Component Models with Relationship matrices"
output: html_document
---

### Example code
To illustrate the usage of the package I simulate a dataset below. The design contains two types of pedigrees, the first are sibling trios with one half sib and two full sibs. The second are two pair of full sibs that are cousins to each other.

```{r sim_relStruct}
# Genetic relationships
str1 <- matrix(c(1, 1/4, 1/4, 1/4, 1, 1/2, 1/4, 1/2, 1), 3, 3)
str2 <- matrix(c(1, 1/2, 1/8, 1/8, 1/2, 1, 1/8, 1/8, 1/8, 1/8, 1, 1/2, 1/8, 1/8, 1/2, 1), 4, 4)

# Genetic relationship matrix
Rgen <- Matrix::bdiag(Matrix::bdiag(replicate(100, str1, simplify = FALSE)),
                      Matrix::bdiag(replicate(100, str2, simplify = FALSE)))
```

Then data can be simulated.
```{r sim_data}
# Number of observations
N <- dim(Rgen)[1]

# Sample genotypic values
a <- MASS::mvrnorm(1, rep(0, N), 2 * Rgen)

# Sample environmental deviations
e <- rnorm(N, 0, sqrt(2))

# Common factor
eta = a + e

# Loadings
l <- matrix(c(1, 0.5, 0.5, 0.8), 4, 1)

# Intercepts
b = c(2, 2, 4, 4)

# Data
Y <- matrix(NA, N, 4, dimnames = list(NULL, paste0("y", 1:4)))
for(i in 1:nrow(Y)) {
 Y[i, ] <- b + l %*% eta[i] + rnorm(4)
}
```

The above code simulates four variables from a common factor model, where the common factor is a function of additive genetic and unique environmental effects. The measurement model for indvidual $i$ is
$$
\boldsymbol{y}_i = \boldsymbol{\beta} + \boldsymbol{\lambda} \eta_i + \boldsymbol{\epsilon}_i.
$$

The structural model is
$$
\eta_i = a_i + e_i.
$$
The distribution of the individual unique factors can be described as
$$
\boldsymbol{\epsilon}_i \sim N(\boldsymbol{0}, \mathbf{\Theta}).
$$
The distribution of the individual unique environmental deviations can be described as
$$
e \sim N(0, \sigma^2_E).
$$
The genotypic values are dependent across individuals. The distribution of the genotypic values across all individuals can be described as
$$
\boldsymbol{a} \sim N(\boldsymbol{0}, \sigma^2_A \mathbf{R}_{gen}).
$$

This model can be specified in **svcmr** as follows.
```{r specify_model}
library(svcmr)

# Relationship matrix for unique environmental deviations
Renv <- Matrix::Diagonal(dim(Rgen)[1])

# Covariates
X <- matrix(1, N, 1)

mod <- svcm(Y,
            # Parameters
            pm(nrow = 4, ncol = 1, labels = paste0("l", 1:4), free = c(F, T, T, T), values = 1, name = "L"),
            pm(nrow = 1, ncol = 1, labels = "Sa1", free = T, values = 1, name = "Sa"),
            pm(nrow = 1, ncol = 1, labels = "Se1", free = T, values = 1, name = "Se"),
            pm(nrow = 4, ncol = 4, labels = paste0("th", 1:16), free = diag(T, 4), values = diag(1, 4), name = "TH"),
            pm(nrow = 4, ncol = 1, labels = paste0("u", 1:4), free = T, values = 0, name = "U"),
            # Variance components
            svc(L %*% Sa %*% t(L), R = Rgen),
            svc(L %*% Se %*% t(L) + TH, R = Renv),
            # Mean components
            mc(U, X = X))
fit <- fitm(mod, se = TRUE, control = list(trace = 6))
summary(fit)
```

Equality constraints can be obtained be using the same labels for several parameters. If we want the variance of the specific factors to be constant we could do it like this
```{r specifiy_model2}
lab_TH <- matrix(NA, 4, 4)
diag(lab_TH) <- "th"
mod2 <- svcm(Y,
            # Parameters
            pm(nrow = 4, ncol = 1, labels = paste0("l", 1:4), free = c(F, T, T, T), values = 1, name = "L"),
            pm(nrow = 1, ncol = 1, labels = "Sa1", free = T, values = 1, name = "Sa"),
            pm(nrow = 1, ncol = 1, labels = "Se1", free = T, values = 1, name = "Se"),
            pm(nrow = 4, ncol = 4, labels = lab_TH, free = diag(T, 4), values = diag(1, 4), name = "TH"),
            pm(nrow = 4, ncol = 1, labels = paste0("u", 1:4), free = T, values = 0, name = "U"),
            # Variance components
            svc(L %*% Sa %*% t(L), R = Rgen),
            svc(L %*% Se %*% t(L) + TH, R = Renv),
            # Mean components
            mc(U, X = X))
fit2 <- fitm(mod2, se = TRUE, control = list(trace = 6))
summary(fit2)

```

The models can the be compared.
```{r compare_models}
anova(fit, fit2)
```

